preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-17 Manuscript/manuscript.Rmd
Working on abstract clarity

**Background** High-throughput chromosome conformation capture technology (Hi-C) has revealed extensive DNA looping and folding into discrete 3D domains. Among these are Topologically Associated Domains (TADs) and chromatin loops, known to play critical roles in cellular processes like gene regulation and cell differentiation. The relatively low resolution of Hi-C data (regions of several kilobases in size) prevents precise mapping of domain boundaries. TAD and loop boundaries have been characterized by the presence of CTCF and members of cohesin complex binding sites, and other genomic annotations. They are typically defined at high resolution (~100bp), suggesting they can inform the precise location of domain boundaries. Several methods have been developed to leverage genome annotation data for predicting TAD/loop boundaries; however, they overlooked key characteristics of the data, such as spatial associations between an annotation and a boundary. Furthermore, the number of regions containing boundaries is much smaller than the rest of the genome, creating a class imbalance problem.

**Methods** To improve the detection of domain boundaries, we developed *preciseTAD*, a random forest-based framework that leverages high-resolution genome annotation data coupled with density-based clustering and scalable partitioning techniques. We investigated several resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) to address class imbalance. We benchmarked our method against the Arrowhead domain caller and a novel chromatin loop prediction algorithm, Peakachu, on multiple cell lines.

**Results** Using different feature engineering and resampling strategies, we found spatial relationships (distance) between boundaries and annotations have the most predictive power. Transcription factor binding sites outperformed other genome annotation types. Random under-sampling significantly improved model performance. Boundaries predicted by preciseTAD were more enriched for CTCF, RAD21, SMC3, and ZNF143 sites and more conserved across cell lines, highlighting their higher biological significance. The model trained in one cell line performs well in predicting boundaries in another cell line.

**Conclusion** Our results outline strategies for predictive modeling of 3D genomic domains using genome annotation data. The precise identification of domain boundaries will improve our understanding of how genomic regulators are associated with the 3D structure of the genome.

**Availability.** Our method is implemented as an R package, *preciseTAD*, available on https://github.com/dozmorovlab/preciseTAD and Bioconductor (submitted).

>>>>

**Background.** High-throughput chromosome conformation capture technology (Hi-C) revealed extensive DNA looping and folding into discrete 3D domains. Among these are Topologically Associated Domains (TADs) and chromatin loops, critical for cellular processes like gene regulation and cell differentiation. The relatively low resolution of Hi-C data (regions of several kilobases in size) prevents precise mapping of domain boundaries. However, the high resolution of genomic annotations associated with boundaries, such as CTCF, members of cohesin complex, suggest they can inform the precise location of domain boundaries. Several methods attempted to leverage genome annotation data for predicting TAD/loop boundaries; however, they overlooked key characteristics of the data, such as spatial associations between an annotation and a boundary, and a much smaller number of boundaries than the rest of the genome (class imbalance).

**Methods.** We developed *preciseTAD*, an optimized random forest model to improve the location of domain boundaries. Trained on high-resolution genome annotation data and boundaries from low-resolution Hi-C data, the model predicts the location of boundaries at base-level resolution. We investigated several feature engineering and resampling techniques (random over- and undersampling, Synthetic Minority Over-sampling TEchnique (SMOTE)) to select the most optimal data characteristics and address class imbalance. Density-based clustering and scalable partitioning techniques were used to identify the precise location of boundary regions. We benchmarked our method against the Arrowhead domain caller and a novel chromatin loop prediction algorithm, Peakachu, on two most annotated cell lines.

**Results.** Spatial relationships (distance) between boundaries and annotations have the most predictive power. Transcription factor binding sites outperformed other genome annotation types. Random under-sampling significantly improved model performance. Boundaries predicted by *preciseTAD* were more enriched for CTCF, RAD21, SMC3, and ZNF143 signal and more conserved across cell lines, highlighting their higher biological significance. The model trained in one cell line performs well in predicting boundaries in another cell line.

**Conclusion.** Our study implements the method and outlines strategies for TAD/loop boundary prediction using genome annotation data. The precise identification of domain boundaries will improve our understanding of how genomic regulators are associated with the 3D structure of the genome.

**Availability.** *preciseTAD* R package is available on https://github.com/dozmorovlab/preciseTAD and Bioconductor (submitted).
